---
title: "Unemployment Heat Mapping"
author: 'Sun Park'
date: "October 2023"
output: pdf_document
---

\newpage

```{r echo=FALSE, warning=FALSE, , message=FALSE}

# Load the required library
library(httr)
library(readr)
library(dplyr)

get_unemployment_data <- function(date_mth = NULL){
    # Set the URL and User-Agent
    url <- 'https://www.bls.gov/web/metro/laucntycur14.txt'
    user_agent <- 'Mozilla/5.0 (X11; Linux x86_64) AppleWebKit/537.36 (KHTML, like Gecko) Chrome/110.0.0.0 Safari/537.36'
    
    # Send a GET request with the specified User-Agent
    response <- GET(url, user_agent(user_agent))
    
    # Extract the content as text
    data <- content(response, as = 'text')
    
    # Split the text into lines
    lines <- unlist(strsplit(data, "\n"))
    # Get data parts
    data <- lines[6:length(lines)]
    
    # Split the data by '|'
    countyemp <- data.frame(do.call(rbind, strsplit(data, "\\|", perl = TRUE)))
    
    # Set column names
    colnames(countyemp) <- c("area_code", "fips_state", "fips_county", "area_title", "period", "labor_force", "employed", "unemployed", "unemployed_rate")
    
    # Drop rows with missing data
    countyemp <- countyemp[-1,]
    # Remove white spaces from all columns
    countyemp <- lapply(countyemp, function(x) gsub("\\s", "", x))
    countyemp <- as.data.frame(countyemp)
    #get the right format of date
    countyemp <- countyemp %>% dplyr::mutate(period=as.Date(paste("01-", period, sep = ""), format = "%d-%b-%y")) 
    
    # Get the FIPS code: Have to add leading zeros to any single digit number and combine them.
    countyemp$fips_county <- formatC(countyemp$fips_county, width = 3, format = "d", flag = "0")
    countyemp$fips_state <- formatC(countyemp$fips_state, width = 2, format = "d", flag = "0")
    # Remove white spaces from all columns
    countyemp <- lapply(countyemp, function(x) gsub("\\s", "", x))
    countyemp <- as.data.frame(countyemp)
    countyemp$fips <- paste(countyemp$fips_state,countyemp$fips_county,sep="")
    #remove any null values in period
    countyemp <- countyemp[!is.na(countyemp$period), ]
  
    
    # Check for date or dates.
    if (!is.null(date_mth)){
        date_mth <- as.Date(paste("01", date_mth, sep = ""), format = '%d %b %Y')
        # Check to see if users date exists in data set.
        dt_exist <- sapply(date_mth, function(x) any(grepl(x, countyemp$period)))
        if(any(dt_exist==FALSE)){
            message("Are you sure your date(s) is published? Please check the BLS release schedule.")
            if(i>Sys.Date()-54){
                stop(message("County-wide statistics are usually published on the third Friday of each month for the previous month."))
            }
            if(i<Sys.Date()-360){
                stop(message("This data set only goes back one year. Make sure your date(s) is correct."))
            }
        }
    }
    
    if (is.null(date_mth)){
        date_mth <- max(countyemp$period)
        date_mth <- as.Date(date_mth, format = '%d %b %Y')
    }
    
    # Put months to loop in list.
    datalist <- purrr::map(date_mth, function(i){
        mth_vals <- subset(countyemp, period==i)
    })
    
    # Rebind.
    df <- do.call(rbind, datalist)
    
    # Correct column data types.
    df %<>% dplyr::mutate(unemployed=as.numeric(gsub(",", "", as.character(unemployed))), employed=as.numeric(gsub(",", "", as.character(employed))),
                          labor_force=as.numeric(gsub(",", "", as.character(labor_force)))) %>% tibble::as_tibble()
    
    return(df)
}
```




```{r 3, results='asis', echo=FALSE, fig.width=8, fig.height=5, warning=FALSE, message=FALSE}

# Load required libraries
library(maps)
library(RColorBrewer)
library(dplyr)  # Load the dplyr package
library(zoo)    # Load the zoo package

# Generate a sequence of dates spanning the last year to the last 90 days
month <- seq.Date(Sys.Date() - 365, Sys.Date() - 90, by = "month")
month <- as.data.frame(month)
month$month <- format(month$month, format = "%B %Y")

# Load county FIPS data
# county.fips: FIPS county codes for US County Map from library(Map)
data("county.fips")

# Select the last and second-to-last months
i <- nrow(month)
j <- nrow(month) - 1

# Get unemployment data for the last two months
unemp1 <- get_unemployment_data(month[i, 1])
unemp2 <- get_unemployment_data(month[j, 1])

# Calculate unemployment rates
unemp1$unemployed_rate <- unemp1$unemployed / unemp1$labor_force
unemp2$unemployed_rate <- unemp2$unemployed / unemp2$labor_force

# Extract year and month from the date
unemp1$period <- as.yearmon(unemp1$period)
unemp2$period <- as.yearmon(unemp2$period)
x <- as.character(unemp1$period[1])
y <- as.character(unemp2$period[1])

# Select relevant columns
unemp1 <- unemp1[, c(10, 8, 9)]
unemp1$fips <- as.numeric(unemp1$fips)
unemp2$fips <- as.numeric(unemp2$fips)
unemp1$unemployed_rate <- as.numeric(unemp1$unemployed_rate) * 100
unemp2$unemployed_rate <- as.numeric(unemp2$unemployed_rate) * 100

# Define color buckets for the first map
colors <- c("#CCE5FF", "#99CCFF", "#66B2FF", "#3399FF", "#0080FF", "#004C99")
unemp1$colorBuckets <- cut(unemp1$unemployed_rate, c(0, 2, 4, 6, 8, 10, 100))
leg.txt <- c("<2%", "2-4%", "4-6%", "6-8%", "8-10%", ">10%")

# Match county FIPS codes to map data for the first map
cnty.fips <- county.fips$fips[match(map("county", plot = FALSE)$names, county.fips$polyname)]
colorsmatched <- unemp1$colorBuckets[match(cnty.fips, unemp1$fips)]

# Create the first heatmap
map("county", col = colors[colorsmatched], fill = TRUE, resolution = 0, lty = 0, projection = "polyconic")
map("state", col = "black", fill = FALSE, add = TRUE, lty = 1, lwd = 0.2, projection = "polyconic")
title(paste("Unemployment by County,", x), sep = " ")
legend("topright", leg.txt, horiz = FALSE, fill = colors, cex = 0.6)


```

```{r 3.2, results='asis', echo=FALSE, fig.width=8, fig.height=5, warning=FALSE, message=FALSE}

# Select relevant columns for the second map
unemp2 <- unemp2[, c(10, 8, 9)]
unemp1 <- unemp1[, -4]

# Merge data from the two months based on FIPS codes
unemp2 <- merge(unemp2, unemp1, by = "fips")

# Calculate the change in unemployment rate
unemp2$delta <- ((unemp2$unemployed_rate.y - unemp2$unemployed_rate.x) / unemp2$unemployed_rate.x) * 100

# Define color buckets for the second map
colors2 <- c("#5aabfc", "#87c3ff", "#b8d9fc", "#dce4ef", "#ffb5bb", "#f75663", "#ff2626")
unemp2$colorBuckets <- as.numeric(cut(unemp2$delta, c(-100, -15, -10, -5, 5, 10, 15, 100)))
leg.txt2 <- c("<(-15%)", "(-15%)-(-10%)", "(-10%)-(-5%)", "(-5%) - 5%", "5%-10%", "10%-15%", ">15%")

# Match county FIPS codes to map data for the second map
cnty.fips2 <- county.fips$fips[match(map("county", plot = FALSE)$names, county.fips$polyname)]
colorsmatched2 <- unemp2$colorBuckets[match(cnty.fips2, unemp2$fips)]

# Create the second heatmap
map("county", col = colors2[colorsmatched2], fill = TRUE, resolution = 0, lty = 0, projection = "polyconic")
map("state", col = "black", fill = FALSE, add = TRUE, lty = 1, lwd = 0.2, projection = "polyconic")
title(paste("Unemployment Change by County,", y, 'to', x), sep = " ")
legend("topright", leg.txt2, horiz = FALSE, fill = colors2, cex = 0.6)

```


*The unemployment rate represents the number of unemployed as a percentage of the labor force. Labor force data are restricted to people 16 years of age and older, who currently reside in 1 of the 50 states or the District of Columbia, who do not reside in institutions (e.g., penal and mental facilities, homes for the aged), and who are not on active duty in the Armed Forces.